---
title: "Data Simulation and Missingness"
author: "Atakan Akgun, Aylin Karapanar"
date: "`r Sys.Date()`"
output: html_notebook
---

# Data Simulation

```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(purrr)
library(MASS) 
library(ggplot2)


set.seed(1234)
```

## Data simualtion with simulated random effects

```{r}
# Define study design
Country <- c("Netherlands", "France", "Spain", "Morocco")
City <- c("Utrecht", "Nijmegen", "Montpellier", "Barcelona", "Rabat", "Casablanca")
Sites <- paste0("Site_", 1:9)
Trap_type <- c("Light_CO2", "HumanOdor_CO2")
Mosquito_species <- c("Culex quinquefasciatus", "Aedes aegypti",
                      "Aedes albopictus", "Anopheles gambiae", "Culex pipiens")
Sampling_dates <- seq.Date(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "month")

design <- expand.grid(
  City = City,
  Site = Sites,
  Sampling_date = Sampling_dates,
  stringsAsFactors = FALSE
)

design$Country <- dplyr::case_when(
  design$City %in% c("Utrecht", "Nijmegen") ~ "Netherlands",
  design$City == "Montpellier" ~ "France",
  design$City == "Barcelona" ~ "Spain",
  design$City %in% c("Rabat", "Casablanca") ~ "Morocco",
  TRUE ~ NA_character_
)

design <- design %>%
  group_by(City) %>%
  mutate(Location.Site = as.integer(factor(Site))) %>%
  ungroup()

# Generate environmental variables

# Add month and season
design <- design %>%
  mutate(
    Month = month(Sampling_date),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5) ~ "Spring",
      Month %in% c(6, 7, 8) ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    )
  )


# Generate green area coverage data
site_characteristics <- design %>%
  distinct(City, Site) %>%
  mutate(
    Green_level = sample(1:3, n(), replace = TRUE),
    Green_area_coverage = case_when(
      Green_level == 1 ~ runif(n(), 0, 33.3),
      Green_level == 2 ~ runif(n(), 33.3, 66.6),
      Green_level == 3 ~ runif(n(), 66.6, 100),
    ),
    Blue_area_coverage = runif(n(), 2, 50)
  )

design <- design %>%
  left_join(site_characteristics, by = c("City", "Site"))

# Temperature - seasonal variation by city
city_temp_params <- data.frame(
  City = City,
  base_temp = c(12, 13, 16, 18, 20, 21),
  temp_amplitude = c(8, 8, 7, 6, 5, 5)
)

design <- design %>%
  left_join(city_temp_params, by = "City") %>%
  mutate(
    Temperature = base_temp + 
                  temp_amplitude * sin(2 * pi * (Month - 3) / 12) +
                  rnorm(n(), 0, 2),
    Temperature = round(pmax(Temperature, 0), 2)
  ) %>%
  dplyr::select(-base_temp, -temp_amplitude)

# Humidity - correlated with temperature
design <- design %>%
  mutate(
    Relative_humidity = 80 - 0.5 * Temperature +
                        rnorm(n(), 0, 5),
    Relative_humidity = round(pmin(pmax(Relative_humidity, 30), 95), 2)
  )

# Wind speed - seasonal variation
design <- design %>%
  mutate(
    Wind_speed_10m = case_when(
      Season == "Winter" ~ runif(n(), 2.5, 7.5),
      Season == "Spring" ~ runif(n(), 2.0, 6.0),
      Season == "Summer" ~ runif(n(), 0.5, 4.5),
      Season == "Fall" ~ runif(n(), 1.5, 5.5)
    ),
    Wind_speed_10m = round(Wind_speed_10m, 2)
  )

# Assign trap types
design$Trap_type <- sample(Trap_type, nrow(design), replace = TRUE)

city_ecology <- data.frame(
  City = City,
  host_base = c(70, 65, 80, 90, 95, 85),  # more hosts in warmer regions
  predator_base = c(50, 55, 60, 75, 70, 65)
)

design <- design %>%
  left_join(city_ecology, by = "City") %>%
  mutate(
    # Seasonal multipliers
    season_multiplier = case_when(
      Season == "Winter" ~ 0.7,
      Season == "Spring" ~ 1.0,
      Season == "Summer" ~ 1.3,
      Season == "Fall"   ~ 0.9
    ),
    
    # Host abundance (correlated with season + city)
    Host_abundance = round(
      host_base * season_multiplier +
      rnorm(n(), 0, 10), 
      0
    ),
    

  # Predator abundance
  Predator_abundance = round(
        predator_base * (0.8 + 0.4 * sin(2*pi*(Month)/12)) + 
        rnorm(n(), 0, 8),
        0
  ),
  
    Predator_abundance = pmax(Predator_abundance, 1), 
    Host_abundance = pmax(Host_abundance, 1)

  ) %>%
  dplyr::select(-host_base, -predator_base)


# Expand to species level
df <- design %>%
  mutate(n_species = sample(2:5, n(), replace = TRUE)) %>%
  rowwise() %>%
  mutate(
    Observed_species = list(sample(Mosquito_species, n_species))
  ) %>%
  ungroup() %>%
  dplyr::select(-n_species) %>%
  tidyr::unnest(Observed_species) %>%
  rename(Mosquito_species = Observed_species)

df <- df %>%
  mutate(
    Host_abundance = rpois(n(), lambda =
      20 +
      8 * Green_level +                  
      5 * (Temperature / 10) +
      rnorm(n(), 0, 3)
    ),
    Predator_abundance = rpois(
      n(),
      lambda =
        40 +
        10 * Green_level +                
        6 * (Relative_humidity / 50) +
        rnorm(n(), 0, 4)
),
    Predator_abundance = pmax(Predator_abundance, 1), 
    Host_abundance = pmax(Host_abundance, 1)
  )


# Random effects for City (intercept and slopes)

n_cities <- length(unique(df$City))

Sigma_city <- matrix(c(
  0.40^2, 0.40 * 0.20 * 0.3,  # Var(intercept), Cov(intercept, slope)
  0.40 * 0.20 * 0.3,  0.20^2  # Cov(slope, intercept), Var(slope)
), nrow = 2, ncol = 2)

# Generate correlated random intercepts and slopes
city_random_effects_matrix <- mvrnorm(
  n = n_cities,
  mu = c(0, 0),  
  Sigma = Sigma_city
)

city_random_effects <- data.frame(
  City = unique(df$City),
  city_intercept = city_random_effects_matrix[, 1],
  city_green_slope = city_random_effects_matrix[, 2]  # Random slope
)

# Random intercepts for Species
species_random_effects <- data.frame(
  Mosquito_species = unique(df$Mosquito_species),
  species_intercept = rnorm(length(unique(df$Mosquito_species)), mean = 0, sd = 0.3)
)

# Random intercepts for Site nested within City
site_random_effects <- df %>%
  distinct(City, Site) %>%
  mutate(
    site_intercept = rnorm(n(), mean = 0, sd = 0.25)
  )

# Join random effects
df <- df %>%
  left_join(city_random_effects, by = "City") %>%
  left_join(species_random_effects, by = "Mosquito_species") %>%
  left_join(site_random_effects, by = c("City", "Site"))


# Standardize predictors
df <- df %>%
  mutate(
    Green_level_scaled = as.numeric(scale(Green_level)),
    Temperature_scaled = as.numeric(scale(Temperature)),
    Humidity_scaled = as.numeric(scale(Relative_humidity)),
    Wind_scaled = as.numeric(scale(Wind_speed_10m)),
    Blue_area_scaled = as.numeric(scale(Blue_area_coverage))
  )

# Generate abundance with random slopes!
df <- df %>%
  mutate(
    log_mu = 2.0 +                                                        
         0.30 * Green_level_scaled +                                  
         0.40 * ifelse(Trap_type == "Light_CO2", 1, 0) +             
         0.20 * Temperature_scaled +                                  
         0.15 * Humidity_scaled +                                 
         -0.10 * Wind_scaled +                                    
         0.05 * Blue_area_scaled +                                    
         0.15 * Green_level_scaled * ifelse(Trap_type == "Light_CO2", 1, 0) +  
         0.10 * log(Host_abundance) +                     
        -0.50 * log(Predator_abundance) +                     
        0.15 * log(Host_abundance) * log(Predator_abundance)  + 


         # random effects
         city_intercept +                                             
         city_green_slope * Green_level_scaled +                      
         species_intercept +                                          
         site_intercept,
    
    mu = exp(log_mu),
    Mosquito_abundance = rnbinom(n(), mu = mu, size = 2)
  )

# Clean up
df <- df %>%
  dplyr::select(-log_mu, -mu, -city_intercept, -city_green_slope, 
                -species_intercept, -site_intercept,
                -Green_level_scaled, -Temperature_scaled, -Humidity_scaled, 
                -Wind_scaled, -Blue_area_scaled)

# Calculate biodiversity index

df <- df %>%
  group_by(City, Site, Sampling_date) %>%
  mutate(
    p_i = Mosquito_abundance / sum(Mosquito_abundance, na.rm = TRUE),
    Biodiversity_index = -sum(p_i * log(p_i + 1e-8), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(-p_i)


write.csv(df, "./Simulated Data/Mock_Mosquito_MixedEffects_w_slopes_Dataset.csv", 
          row.names = FALSE)

```


# Expected Beta Coefficients

When the raw values are used to fit the models, these will be the expected results to be recovered by the model. 
```{r}
# Load your simulated data
df <- read.csv("./Simulated Data/Mock_Mosquito_MixedEffects_w_slopes_Dataset.csv")

# Simulated coefficients (on standardized scale)
beta_green_std <- 0.30
beta_trap_std <- 0.40
beta_temp_std <- 0.20
beta_humidity_std <- 0.15
beta_wind_std <- -0.10
beta_blue_std <- 0.05
beta_interaction_std <- 0.15

```
```{r}
# Calculate SDs for each predictor
sd_green <- sd(df$Green_area_coverage)
sd_temp <- sd(df$Temperature)
sd_humidity <- sd(df$Relative_humidity)
sd_wind <- sd(df$Wind_speed_10m)
sd_blue <- sd(df$Blue_area_coverage)


# Convert to raw scale
beta_green_raw <- beta_green_std / sd_green
beta_trap_raw <- beta_trap_std  
beta_temp_raw <- beta_temp_std / sd_temp
beta_humidity_raw <- beta_humidity_std / sd_humidity
beta_wind_raw <- beta_wind_std / sd_wind
beta_blue_raw <- beta_blue_std / sd_blue
beta_interaction_raw <- beta_interaction_std / sd_green  

# Display expected raw coefficients
cat("Expected raw coefficients:\n")
cat("Green coverage:", round(beta_green_raw, 4), "\n")
cat("Trap type:", round(beta_trap_raw, 4), "\n")
cat("Temperature:", round(beta_temp_raw, 4), "\n")
cat("Humidity:", round(beta_humidity_raw, 4), "\n")
cat("Wind speed:", round(beta_wind_raw, 4), "\n")
cat("Blue area:", round(beta_blue_raw, 4), "\n")
cat("Green × Trap:", round(beta_interaction_raw, 4), "\n")
```
We need to make linear approximation to find the expected Beta coefficients when raw values are used. Starting with the log-linear relationship:
\begin{equation}
\log(\mu) = \beta \log(X)
\end{equation}

Taking the derivative with respect to $X$:
\begin{equation}
\frac{d[\log(\mu)]}{dX} = \beta \cdot \frac{d[\log(X)]}{dX} = \beta \cdot \frac{1}{X} = \frac{\beta}{X}
\end{equation}

For the raw-linear model:
\begin{equation}
\frac{d[\log(\mu)]}{dX} = \gamma
\end{equation}

To approximate $\gamma$ using the log-linear relationship, evaluate at the mean of $X$:
\begin{equation}
\gamma \approx \frac{\beta}{\bar{X}}
\end{equation}

where $\bar{X} = \text{mean}(X)$.
```{r}
# Calculate means
mean_host <- mean(df$Host_abundance)
mean_predator <- mean(df$Predator_abundance)

# Standardized coefficients (on log scale)
beta_host_log_std <- 0.10
beta_predator_log_std <- -0.50

sd_log_host <- sd(log(df$Host_abundance))
sd_log_predator <- sd(log(df$Predator_abundance))

beta_host_log_raw <- beta_host_log_std / sd_log_host
beta_predator_log_raw <- beta_predator_log_std / sd_log_predator

beta_host_raw_approx <- beta_host_log_raw / mean_host
beta_predator_raw_approx <- beta_predator_log_raw / mean_predator

cat("  Host (raw):", beta_host_raw_approx, "\n")
cat("  Predator (raw):", beta_predator_raw_approx, "\n\n")

cat("  Mean Host:", mean_host, "\n")
cat("  Mean Predator:", mean_predator, "\n")
```


